### 2019-06-06 dmtcs_submission.d/abstract.tex
ssr

Many researchers have studied symmetry properties of 
various Boolean functions. 
A class of Boolean functions, 
called \textbf{nested canalyzing functions} (NCFs),
has been used to model certain biological phenomena.
The purpose of this brief contribution is to bring out 
some interesting relationships between NCFs, symmetric 
Boolean functions and a generalization of symmetric Boolean functions,
which we call $r$-symmetric functions (where $r$ is the symmetry level).
Using a normalized representation for NCFs, we develop a 
characterization of when two variables of an NCF are symmetric.
Using this characterization, we show 
that the symmetry level of an NCF $f$
can be easily computed given a standard representation of $f$.
%% (In contrast, we prove that even approximating the symmetry level of
%% a general Boolean function to within any factor $\geq 1$ is \cnp-hard.) 
We also present an efficient algorithm for testing whether 
a given $r$-symmetric function is an NCF.
Further, we show that for any NCF $f$ with $n$ variables, the notion of
strong asymmetry considered in the literature is equivalent to
the property that $f$ is $n$-symmetric. 
We use this result to derive a closed form expression for the
number of $n$-variable Boolean functions 
that are NCFs and strongly asymmetric.
We also identify all the Boolean functions that are NCFs 
and symmetric.

>>>>

Many researchers have studied symmetry properties of 
various Boolean functions. 
A class of Boolean functions, 
called \textbf{nested canalyzing functions} (NCFs),
has been used to model certain biological phenomena.
We identify some interesting relationships between NCFs, symmetric 
Boolean functions and a generalization of symmetric Boolean functions,
which we call $r$-symmetric functions (where $r$ is the symmetry level).
Using a normalized representation for NCFs, we develop a 
characterization of when two variables of an NCF are symmetric.
Using this characterization, we show 
that the symmetry level of an NCF $f$
can be easily computed given a standard representation of $f$.
%% (In contrast, we prove that even approximating the symmetry level of
%% a general Boolean function to within any factor $\geq 1$ is \cnp-hard.) 
We also present an efficient algorithm for testing whether 
a given $r$-symmetric function is an NCF.
Further, we show that for any NCF $f$ with $n$ variables, the notion of
strong asymmetry considered in the literature is equivalent to
the property that $f$ is $n$-symmetric. 
We use this result to derive a closed form expression for the
number of $n$-variable Boolean functions 
that are NCFs and strongly asymmetric.
We also identify all the Boolean functions that are NCFs 
and symmetric.

### 2019-06-06 dmtcs_submission.d/intro.tex
ssr

Rsearch on the symmetry properties of Boolean
functions has received a lot of attention from the
logic design and automated synthesis communities. 
(Additional discussion on this is provided in Section~\ref{sse:related}.)
In this paper, we focus on the symmetry properties of
a class of Boolean functions (called \textbf{nested canalyzing functions})
which are used to model biological phenomena.
To define this class, we start with a simpler
class of functions, called \textbf{canalyzing} Boolean functions.
This class, introduced in \cite{Kauffman-1969}, is defined as follows.

@@

\noindent 
where $x_{\pi(i)}$ is the \textbf{canalyzing variable} that is
\textbf{tested} in line $i$, 
and $a_i$ and $b_i$ are respectively the canalyzing and 
canalyzed values in line $i$,~ $1 \leq i \leq n$.
Each such line is called a \textbf{rule}.
When none of the conditions ``$x_{\pi(i)} ~=~ a_i$" 
is satisfied, we have line $n+1$ with the ``Default" rule
for which the canalyzed value is~ $\overline{b_n}$:

@@

In the literature (see e.g., \cite{Crama-Hammer-2011,HT-2016,Toth-etal-1977}),
a 1-symmetric function $f$ is referred to simply
as a \textbf{symmetric} function or 
as a \textbf{totally symmetric function} \cite{Biswas-1970,BS-1968}.
Thus, in any symmetric function, each pair of variables is symmetric.
As a simple example, the function $f_2(x_1, x_2, x_3)$ = 
$x_1 \oplus x_2 \oplus x_3$ (defined above) is symmetric.
If $f$ is a symmetric function, then for any
input $(a_1, a_2, \ldots, a_n)$ to $f$, where $a_i \in \{0,1\}$ for
$1 \leq i \leq n$, and any permutation $\pi$ of $\{1, 2, \ldots, n\}$,
$f(a_1, a_2, \ldots, a_n)$ = $f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.
For $r \geq 1$, the class of $r$-symmetric functions has been
studied in the literature on discrete dynamical systems (see e.g., 
\cite{Barrett-etal-2007,Rosenkrantz-etal-2015,MR-2007}).

Other forms of symmetry which can be more general than the permutations 
corresponding to symmetry groups have also been considered (see e.g.,
\cite{Maurer-2015,KS-2000}).
A Boolean function $f(x_1, x_2, \ldots, x_n)$
is \textbf{strongly asymmetric}
if for any permutation $\pi$ of $\{1, 2, \ldots, n\}$
\emph{except} the identity permutation,
there exists an input $(a_1, a_2, \ldots,  a_n)$
to $f$ such that $f(a_1, a_2, \ldots, a_n)$ $\neq$
$f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.
In general, there are functions with $n$ variables that are properly
$n$-symmetric, but not strongly asymmetric \cite{KS-2000}.
However, in the case of NCFs with $n$ variables, we will show 
(see Section~\ref{sec:ncf_and_symmetry}) that the notion of 
strong asymmetry coincides with that 
of being properly $n$-symmetric.

@@

The usefulness of symmetry properties in synthesizing 
combinational logic functions was first noted 
by Shannon \cite{Shannon-1938}.
Over the years, a considerable amount of research on detecting 
and exploiting symmetry properties 
of Boolean functions has been reported in the literature.
For example, references \cite{Biswas-1970,TM-1996} present methods 
to determine whether a given Boolean function is symmetric.
Born and Scidmore \cite{BS-1968} show how certain Boolean functions
can be converted into symmetric functions by introducing additional
variables. 
%% (e.g., \cite{BS-1968,Biswas-1970,TM-1996,KS-2000,Darga-etal-2008,Maurer-2015}).
Several researchers have developed techniques for exploiting symmetries
for automated logic synthesis 
(e.g., \cite{AP-2008,Hu-etal-2008,KS-2000,Darga-etal-2008}).
Maurer \cite{Maurer-2015} provides a thorough discussion of known
symmetry detection algorithms and presents a new universal algorithm 
for detecting any form of permutation-based symmetry in Boolean functions.

The term \textbf{canalization}, coined by
Waddington \cite{Waddington-1942}, is generally used to describe
the stability of a biological system with changes
in external conditions.
In 1969, Kauffman \cite{Kauffman-1969} introduced canalyzing Boolean functions
to explain the stability of gene regulatory networks.
The subclass of NCFs was
introduced later by Kauffman et al. \cite{Kauffman-etal-2003} 
to facilitate a rigorous analysis of the Boolean network model
for gene regulatory networks.
It is known that the class of NCFs coincides with that
of unate cascade Boolean functions \cite{Jarrah-etal-2007}.
In the literature on computational learning theory,
NCFs are referred to as $1$-\textbf{decision lists} \cite{KV-1994}.
Many researchers have pointed out the usefulness of NCFs 
in modeling biological phenomena 
(e.g., \cite{Layne-2011,
Layne-etal-2012,Li-etal-2011,Li-etal-2012,Li-etal-2013}).
Properties of NCFs such as sensitivity and 
stability have also been studied in the 
literature \cite{Kauffman-etal-2004, Layne-2011,Layne-etal-2012,
Li-etal-2011,Li-etal-2013,Klotz-etal-2013, Stearns-etal-2018}.

>>>>

Research on the symmetry properties of Boolean
functions has received a lot of attention from the
logic design and automated synthesis communities. 
(Additional discussion on this is provided in Section~\ref{sse:related}.)
In this paper, we focus on the symmetry properties of
a class of Boolean functions (called \textbf{nested canalyzing functions})
which are used to model biological phenomena.
To define this class, we start with a simpler
class of functions, called \textbf{canalyzing} Boolean functions.
This class, introduced in \cite{Kauffman-1969}, is defined as follows.

@@

\noindent 
where $x_{\pi(i)}$ is the \textbf{canalyzing variable} that is
\textbf{tested} in line $i$, 
and $a_i$ and $b_i$ are respectively the \emph{canalyzing} and 
\emph{canalyzed} values in line $i$,~ $1 \leq i \leq n$.
Each such line is called a \textbf{rule}.
When none of the conditions ``$x_{\pi(i)} ~=~ a_i$" 
is satisfied, we have line $n+1$ with the ``Default" rule
for which the canalyzed value is~ $\overline{b_n}$:

@@

In the literature (see e.g., \cite{Crama-Hammer-2011,HT-2016,Toth-etal-1977}),
a 1-symmetric function $f$ is referred to simply
as a \textbf{symmetric} function or 
as a \textbf{totally symmetric function} \citep{Biswas-1970,BS-1968}.
Thus, in any symmetric function, each pair of variables is symmetric.
As a simple example, the function $f_2(x_1, x_2, x_3)$ = 
$x_1 \oplus x_2 \oplus x_3$ (defined above) is symmetric.
If $f$ is a symmetric function, then for any
input $(a_1, a_2, \ldots, a_n)$ to $f$, where $a_i \in \{0,1\}$ for
$1 \leq i \leq n$, and any permutation $\pi$ of $\{1, 2, \ldots, n\}$,
$f(a_1, a_2, \ldots, a_n)$ = $f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.
For $r \geq 1$, the class of $r$-symmetric functions has been
studied in the literature on discrete dynamical systems (see e.g., 
\cite{Barrett-etal-2007,Rosenkrantz-etal-2015,MR-2007}).

Other forms of symmetry which can be more general than the permutations 
corresponding to symmetry groups have also been considered (see e.g.,
\cite{Maurer-2015,KS-2000}).
A Boolean function $f(x_1, x_2, \ldots, x_n)$
is \textbf{strongly asymmetric}
if for any permutation $\pi$ of $\{1, 2, \ldots, n\}$
\emph{except} the identity permutation,
there exists an input $(a_1, a_2, \ldots,  a_n)$
to $f$ such that $f(a_1, a_2, \ldots, a_n)$ $\neq$
$f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.
In general, there are functions with $n$ variables that are properly
$n$-symmetric, but not strongly asymmetric \citep{KS-2000}.
However, in the case of NCFs with $n$ variables, we will show 
(see Section~\ref{sec:ncf_and_symmetry}) that the notion of 
strong asymmetry coincides with that 
of being properly $n$-symmetric.

@@

The usefulness of symmetry properties in synthesizing 
combinational logic functions was first noted 
by \citet{Shannon-1938}.
Over the years, a considerable amount of research on detecting 
and exploiting symmetry properties 
of Boolean functions has been reported in the literature.
For example, \citet{Biswas-1970} and \citet{TM-1996} present methods 
to determine whether a given Boolean function is symmetric.
\citet*{BS-1968} show how certain Boolean functions
can be converted into symmetric functions by introducing additional
variables. 
%% (e.g., \cite{BS-1968,Biswas-1970,TM-1996,KS-2000,Darga-etal-2008,Maurer-2015}).
Several researchers have developed techniques for exploiting symmetries
for automated logic synthesis 
(e.g., \cite{AP-2008,Hu-etal-2008,KS-2000,Darga-etal-2008}).
\citet{Maurer-2015} provides a thorough discussion of known
symmetry detection algorithms and presents a new universal algorithm 
for detecting any form of permutation-based symmetry in Boolean functions.

The term \textbf{canalization}, coined by
\citet{Waddington-1942}, is generally used to describe
the stability of a biological system with changes
in external conditions.
\citet{Kauffman-1969} introduced canalyzing Boolean functions
to explain the stability of gene regulatory networks.
The subclass of NCFs was
introduced later by \citet{Kauffman-etal-2003} 
to facilitate a rigorous analysis of the Boolean network model
for gene regulatory networks.
It is known that the class of NCFs coincides with that
of unate cascade Boolean functions \citep{Jarrah-etal-2007}.
In the literature on computational learning theory,
NCFs are referred to as $1$-\textbf{decision lists} \citep{KV-1994}.
Many researchers have pointed out the usefulness of NCFs 
in modeling biological phenomena 
(e.g., \cite{Layne-2011,
Layne-etal-2012,Li-etal-2011,Li-etal-2012,Li-etal-2013}).
Properties of NCFs such as sensitivity and 
stability have also been studied in the 
literature (e.g., \cite{Kauffman-etal-2004, Layne-2011,Layne-etal-2012,
Li-etal-2011,Li-etal-2013,Klotz-etal-2013, Stearns-etal-2018}).
